Parametricity Consequences of GADTs Representations

### 2021-02-28 20-gadts/GADT-story.tex
finish draft of Section 4

To see that the Church encoding of a GADT and its primitive
representation need not be the same, consider again the GADT
$\mathsf{G\,a}$ defined in~\eqref{eq:G}. Despite its simplicity, this
GADT serves as an informative case study highlighting the differences
--- even if we only consider the data elements a GADT contains ---
between a GADT's Church encoding its representation as a primitive
data type.

\begin{example}\label{ex:g1}
Syntactically, the GADT $\mathsf{G\,a}$ defined in~\eqref{eq:G}
comprises a single data element, namely $\mathsf{c :: G \, 1}$. This
is captured by $\mathsf{G}$'s Church encoding
\[\mathsf{G\,a} = \mathsf{Equal\,1\,a}\]
from~\cite{atk12}, so $\mathsf{G}$'s effect is simply to test its
argument for equality against the unit type
$\mathsf{1}$. {\color{blue} Use same notation for unit type as in
  intro.}
%Recalling from Section~\ref{sec:CEs} that the
The interpretation of $\mathsf{G\,a}$ represented as the Chirch
encoding $\mathsf{Equal\,1\,a}$ is thus the function whose value is a
singleton set when the interpretation of $\mathsf{a}$ is $1$ and the
empty set otherwise. This perfectly mirrors $\mathsf{G}$'s syntax,
which indeed delivers just one data element when $\mathsf{a}$ is
$\mathsf{1}$ and no elements otherwise.

We now want to compute the interpretation of $\mathsf{G}$'s primitive
representation. The type of $\mathsf{G}$'s unique constructor
$\mathsf{c :: G\,1}$ can be equivalently expressed as $\mathsf{c :: 1
  \to G\,1}$ or, using~\eqref{eq:nat-transfs}, as $\mathsf{c ::
  (Lan_{\lambda *. 1}\,\lambda *. 1)\,a \to G\, a}$, where
$\mathsf{\lambda *. 1}$ is the syntactic reflection of the constantly
$1$-valued functor from the category with a single object $*$ to
$\mathit{Set}$. We can therefore represent $\mathsf{G}$ as
\[\mathsf{G\,a} = (\mathsf{\mu \phi. \lambda b.
 (Lan_{\lambda *. 1} \,\lambda *.  1)\,b)\,a}\] The interpretation of
$\mathsf{G\, a}$ is therefore just the application to $\mathsf{a}$ of
the fixpoint of the interpretation of the body $\mathsf{(Lan_{\lambda
    *. 1} \,\lambda *. 1)\,b}$ of the syntactic fixpoint.  But since
the recursion variable $\mathsf{\phi}$ does not appear in this body,
the interpretation of the fixpoint is just the interpretation of the
body itself.  The interpretation of $\mathsf{G\,a}$ is therefore
$(\mathit{Lan}_{\lambda *. 1} \lambda *. 1)\, A$, where $A$ interprets
$\mathsf{a}$.
%and $1$ is the {\color{blue} constantly $1$-valued
%  functor $\ast \to \set$ from the category with a single object
%  $\ast$ to $\set$.}
It turns out, however, that, for any set $A$, $(\mathit{Lan}_{\lambda
  *. 1} \lambda *. 1)\,A$ is, in fact, exactly $A$. Indeed,
Proposition~7.1 of~\cite{blw03} gives that $(\mathit{Lan}_{\lambda
  *. 1} \lambda *. 1)\,A$ can be computed as the set of all functions
$a : 1 \to A$ modulo the smallest equivalence relation $\sim$ such
that, for any $f : 1 \to 1$, we have $a \sim a \circ f$.  But this set
is isomorphic to $A$, because the only function $f : 1 \to 1$ is the
identity function.

@@

{\color{red} STOPPED HERE}

Non-uniform nature of GADTs renders build, therefore short cut, for
GADTs impossible. Can't even formulate! Of course, there could be
another way to prove equivalence.  But, in fact, CEs and primitive
representations are not equivalent for arbitrary GADTs, as the
following example shows.
{\color{blue} Our favorite GADT isn't a hereditarily
  polynomial nested type.}

Forced to choose, a programmer would likely find the idea that a GADT
contains data that are not specified by its syntax more than a little
disturbing. What, they might ask, should a data type contain other
than data that are constructed using its data constuctors? Why should
a GADT contain ``hidden'' elements that are only accessible by
mapping? On the other hand, a semanticist would likely find the
primitive representation of GADTs more reasonable. Indeed, they would
likely find the nonfunctorial nature of the Church encoding unnerving
at best. After all, they would likely argue, the data in a GADT
shouldn't disappear just because you map a function over it! The fact
that this happens when GADTs are represented by their Church encodings
actually highlights how GADTs {\em do not} generalize the essential,
container-ish nature of ADTs at all. A semanticist might therefore
conclude that GADTs are seriously misnamed. They might even go on to
wonder: in what sense are GADTs data types at all?

The disappointment that GADTs don't validate any such equivalence is
further compounded by the fact that the the two representations have
very different parametricity properties. We explore the differences in
parametricity results for the two representations in the next section.

>>>>

To see that the Church encoding of a GADT and its primitive
representation need not be the same, consider again the GADT
$\mathsf{G}$ defined in~\eqref{eq:G}. Despite its simplicity, this
GADT serves as an informative case study highlighting the differences
between a GADT's Church encoding its primitive representation --- even
if we consider only the data elements it contains, and ignore whether
or not it supports a $\mathsf{map}$ function.

\begin{example}\label{ex:g1}
Syntactically, the GADT $\mathsf{G}$ defined in~\eqref{eq:G} comprises
a single data element, namely $\mathsf{c :: G \, 1}$. This is captured
by $\mathsf{G}$'s Church encoding
\[\mathsf{G\,a} = \mathsf{Equal\,1\,a}\]
from~\cite{atk12}. As its Church encoding makes clear, $\mathsf{G}$'s
effect is simply to test its argument for equality against the unit
type $\mathsf{1}$. {\color{blue} Use same notation for unit type as in
  intro.} The interpretation of the Church encoding of $\mathsf{G}$
is thus, in accordance with Section~\ref{sec:CEs}, the function whose
value is a singleton set when the interpretation of $\mathsf{a}$ is
$1$ and the empty set otherwise. This perfectly accords with
$\mathsf{G}$'s syntax, which indeed delivers just one data element
when $\mathsf{a}$ is $\mathsf{1}$ and no elements otherwise.

To compute the interpretation of $\mathsf{G}$'s primitive
representation we first note that the type of $\mathsf{G}$'s solitary
constructor $\mathsf{c :: G\,1}$ can be equivalently expressed as
$\mathsf{c :: 1 \to G\,1}$ or, using~\eqref{eq:nat-transfs}, as
$\mathsf{c :: (Lan_{\lambda *. 1}\,\lambda *. 1)\,a \to G\, a}$, where
$\mathsf{\lambda *. 1}$ is the syntactic reflection of the constantly
$1$-valued functor from the category with a single object $*$ to
$\mathit{Set}$. We can therefore represent $\mathsf{G}$ as
\[\mathsf{G\,a} = \mathsf{(\mu \phi. \lambda b.
 (Lan_{\lambda *. 1} \,\lambda *.  1)\,b)\,a}\] The interpretation of
$\mathsf{G}$ is therefore the application to $\mathsf{a}$ of the
fixpoint of the interpretation of the body $\mathsf{\lambda
  b. (Lan_{\lambda *. 1} \,\lambda *. 1)\,b}$ of the syntactic
fixpoint $\mathsf{\mu \phi. \lambda b.  (Lan_{\lambda *. 1} \,\lambda
  *.  1)\,b}$. But since the recursion variable $\mathsf{\phi}$ does
not appear in this body, the interpretation of the fixpoint is just
the interpretation of the body itself.  The interpretation of
$\mathsf{G}$ is therefore $(\mathit{Lan}_{\lambda *. 1} \lambda
*. 1)\, A$, where $A$ interprets $\mathsf{a}$.
%and $1$ is the {\color{blue} constantly $1$-valued
%  functor $\ast \to \set$ from the category with a single object
%  $\ast$ to $\set$.}
It turns out, however, that, for any set $A$, $(\mathit{Lan}_{\lambda
  *. 1} \lambda *. 1)\,A$ is, in fact, exactly $A$. Indeed,
Proposition~7.1 of~\cite{blw03} gives that $(\mathit{Lan}_{\lambda
  *. 1} \lambda *. 1)\,A$ can be computed as the set of all functions
$a : 1 \to A$ modulo the smallest equivalence relation $\sim$ such
that, for any $f : 1 \to 1$, we have $a \sim a \circ f$.  But this set
is isomorphic to $A$, because the only function $f : 1 \to 1$ is the
identity function.

@@

Forced to choose, a programmer would likely find the idea that a GADT
contains data not specified by its syntax more than a little
disturbing. What, they might ask, should a data type contain other
than data that are constructed using its data constuctors? Why should
a GADT contain ``hidden'' elements that are not specified by the
GADT's syntax and are only accessible via applications of
$\mathsf{map}$? A semanticist, on the other hand, would likely find
the primitive representation of GADTs entirely reasonable. Indeed,
they would likely find the nonfunctorial nature of a GADT's Church
encoding unnerving at best. After all, they would likely argue, the
data in a GADT shouldn't disappear just because you map a function
over it! The fact that this happens when GADTs are represented by
their Church encodings actually highlights how GADTs {\em do not}
generalize the essential, container-ish nature of ADTs at all. A
semanticist might therefore conclude that GADTs are seriously
misnamed. They might even go on to wonder: in what sense do GADTs
really deserve to be considered {\em data types} at all?

The reader familiar with short cut fusion will quickly see that the
non-uniform nature of GADTs that are not nested types entails that
short cut fusion cannot be validated for their Church encodings. In
fact, it is not even possible to define a $\mathsf{build}$ function,
which is a crucial ingredient in short cut fusion, for the Church
encoding of a GADT that is not a nested type. But the fact that not
all of the elements in the functorial completion of a GADT that is not
a nested type are directly representable as applications of data
constructors means that it is also impossible to define a
$\mathsf{build}$ function for the primitive representation of such a
GADT.  This is not asurorising, however: if short cut fusion were to
hold for a GADT then its Church encoding and primitive representation
would necessarily be the equivalent.

In addition to not being equivalent, the two representations of GADTs
have very different implications for parametricity for GADTs. We
explore the differences in parametricity results for the two
representations in the next section.
